Fast and optimal centroiding of faint stars

### 2015-07-16 ms/figure.tex
Update figure.tex

\begin{abstract}
One of the most demanding tasks in astronomical image processing---in terms of precision---is the centroiding of stars. 
Upcoming large surveys are going to take images of billions of point sources, 
including many faint stars, with short exposure times. Real-time estimation of the
centroids of stars is crucial for real-time PSF estimation, and maximal precision is required
for measurements of proper motion. In this work, we aim to compare the performance of
various centroiding methods, when they are applied
to relatively low signal-to-noise ratio unsaturated stars.
In order to investigate how information-preserving these techniques are, we compare
the root-mean-squared-error with the fundamental Cramer-Rao
lower bound assuming zero-mean constant Gaussian noise. We discuss two general circumstances in centroiding of faint stars: (i) when we have a good estimate
of the PSF, (ii) when we do not know the PSF. In the case that we know the PSF, 
we show that a fast polynomial centroiding after smoothing the image by the PSF can be 
as accurate as full PSF profile fitting. In the case that we do not
know the PSF, we demonstrate that although polynomial centroiding is not as
accurate as PSF profile fitting, it comes close to saturating the Cramer-Rao lower bound
in a wide range of conditions. We also show that the center-of-light never comes close to 
saturating the bound, and thus it does not deliver reliable estimates of centroids.

@@

In this paper, we study how optimal various techniques are at centroiding 
faint unsaturated stars. That is, we apply a number of centroiding methods 
to a large number of simulated faint stars with different signal-to-noise ratio and
 size realizations. Error from star centroiding methods, will always have 
a theoretically-set lower bound, known as the Cramer-Rao lower bound
(hereafter denoted by CRLB) which has an inverse relation with the
signal-to-noise-ratio of stars.

@@

This paper is structured as follows. In section \ref{sec:CRLB},
we discuss the Cramer-Rao lower bound, and we derive
an analytic expression for the lower bound on centroiding error
of the simulated data, and also PSF profile fitting. 
In section \ref{sec:method} we give a brief overview of approximate 
centroiding methods used in our investigation.
In section \ref{sec:data} we discuss the Cramer-Rao lower bound satuaration
tests and their corresponding simulated data.
In section \ref{sec:result}, we compare the performances of the methods
discussed in \ref{sec:method}, with the CRLB derived in \ref{sec:data}. Finally,
 we discuss and conclude in \ref{sec:discussion}.               
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Cramer-Rao lower Bound %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Cramer-Rao lower bound on centroiding error}\label{sec:CRLB}

In order to test the accuracy of various centroiding methods, we need to to compare their 
performances against each other at saturating the Cramer-Rao lower bound on centroiding error. 
The closer an estimator is to saturating 
the CRLB, the more information about the quantity that we 
need to estimate is preserved. A measure for closeness of an estimator to 
saturating the CRLB is comparison of the root-mean-squared-error of that estimator 
with the CRLB.

The CRLB is given by square-root of the inverse of 
the Fisher information matrix $\mathcal{F}$. Thus, in order to find the CRLB,
 it is sufficient to compute the Fisher matrix. Let us assume that there are $M$ observables $\{f_{1}, ... , f_{M}\}$, each
related to $B$ model parameters $\{\theta_{1} , ... , \theta_{B}\}$
\beq
f_{m} = f_{m}(\theta_{1} , ... , \theta_{B}).
\eeq

@@

We examine fitting an exact
 PSF profile to the stars. That is, in our Cramer-Rao bound saturation tests, we find the best
 estimates of flux and centroid by optimizing the $\chi^{2}$.
 We expect this method to perform better in determing the centroids
 of stars, and deliver RMSE very close to Cramer-Rao bound. In the next section,
 we briefly discuss the centroiding methods used in our Cramer-Rao bound saturating tests.

@@

In terms of saturating the Cramer-Rao lower bound, we expect this simple 
center of light centroiding to perform worse than all other methods mentioned in 
this section. Hereafter, we call this method $7\times7$ moment centroiding.

@@

An optimal stellar centroiding algorithm must saturate the fundamental Cramer-Rao lower 
bound. That is, in all ranges of background noise level, size, radial light profile,
and shape, it must preserve information about the centroids of stars. In practice however,
this is only achievable when we have a reasonably good estimate of the PSF. Since, we do not always 
know the exact PSF profile, we make use of various approximate centroiding algorithms. In this work, we
studied how different methods come close to saturating the CRLB for relatively low signal-to-noise ratio
unsaturated stars.

We focused on examples from two classes of centroiding algorithms. The first class contains fast and approximate
methods that do not require any knowledge of the PSF at the positions of stars. Of methods that belong to this class,
we consider centroiding stars based on fitting a second-order
polynomial to a 3$\times$3 patch of star images smoothed by a Gaussian kernel of fixed width, and finding the center of light 
of a 7$\times$7 patch around the brightest pixel of the star.
The second class of centroiding algorithms contains methods that require knowledge of the PSF (or having a good estimate
of the PSF) at the positions of stars. We considered two examples from this class. The first example
is the matched-filter polynomial centroiding, and the
second example is the full PSF profile fitting. In terms of saturating the Cramer-Rao bound, we compared 
the performances of these methods against each other.

@@

On the other hand, the RMSE of centroid estimates of 
default polynomial centroiding are much closer to saturating the CRLB in all ranges of signal-to-noise ratio.
The performance of this method however, is limited by
two important factors: \emph{(i)} signal-to-noise ratio, and \emph{(ii)} PSF size.
The default polynomial centroiding technique only takes advantage of the information
 contained in a 3$\times$3 patch centered on the brighest pixel of the smoothed image which is only
 well-sampled when the FWHM of the simulated image of star
 matches that of the smoothing kernel. Thus, when we apply 
this method to find the centroids of stars with larger FWHM, certain amount of
information (encoded in the Cramer-Rao lower bound) is lost, and therefore, the RMSE
of these methods deviates from the CRLB. This deviation becomes even larger at lower
signal-to-noise ratios. Besides, the performance of this method slightly degrades
in the case of undersampled stars (with FWHM close two 2 pixels). 
Presence of noise is another limitting factor.
Although this method is able to get very close to saturating the CRLB in a wide range of
signal-to-noise ratios, it is not reliable in the case of centroiding extremely faint stars.
 This is partly due to the fact that in the presence of noise, 
the brightest pixel of image does not necessarily contain the centroid of stars even after smoothing 
the image.

@@

\bibitem[Bertin \& Arnouts (1996)]{sex} Bertin, E., Arnouts, S. \ 1996  Astronomy and Astrophysics Supplement Series, 117, 393-404
\bibitem[Lupton etal. (2001)]{sdss} Lupton, R. etal. \ 2001  arXiv: astro-ph/0101420
\bibitem[Schechter etal. (1993)]{dophot} Schechter, P. L., Mateo, M., Saha, A. \ 1993 \pasp, 1342-1353
\bibitem[Stetson (1987)]{daophot} Stetson, P. B. \ 1987, \pasp, 191-222
\bibitem[Trujillo etal. (2001)]{moffat} Trujillo, I., et al. \ 2001 \mnras, 977-985

>>>>

\begin{abstract}
One of the most demanding tasks in astronomical image processing---in terms of precision---is the centroiding of stars. 
Upcoming large surveys are going to take images of billions of point sources, 
including many faint stars, with short exposure times. Real-time estimation of the
centroids of stars is crucial for real-time PSF estimation, and maximal precision is required
for measurements of proper motion. In this work, we aim to compare the performance of
various centroiding methods, when they are applied
to relatively low signal-to-noise ratio unsaturated stars.
In order to investigate how information-preserving these techniques are, we compare
the root-mean-squared-error with the fundamental Cram\'{e}r-Rao
lower bound assuming zero-mean constant Gaussian noise. We discuss two general circumstances in centroiding of faint stars: (i) when we have a good estimate
of the PSF, (ii) when we do not know the PSF. In the case that we know the PSF, 
we show that a fast polynomial centroiding after smoothing the image by the PSF can be 
as accurate as full PSF profile fitting. In the case that we do not
know the PSF, we demonstrate that although polynomial centroiding is not as
accurate as PSF profile fitting, it comes close to saturating the Cram\'{e}r-Rao lower bound
in a wide range of conditions. We also show that the center-of-light never comes close to 
saturating the bound, and thus it does not deliver reliable estimates of centroids.

@@

In this paper, we study how optimal various techniques are at centroiding 
faint unsaturated stars. That is, we apply a number of centroiding methods 
to a large number of simulated faint stars, assuming uncorrelated Gaussian noise, 
with different signal-to-noise ratio and size realizations. Error from star 
centroiding methods, will always have a theoretically-set lower bound, 
known as the Cram\'{e}r-Rao lower bound (hereafter denoted by CRLB) which 
has an inverse relation with the signal-to-noise-ratio of stars.

@@

This paper is structured as follows. In section \ref{sec:CRLB},
we discuss the Cram\'{e}r-Rao lower bound, and we derive
an analytic expression for the lower bound on centroiding error
of the simulated data, and also PSF profile fitting. 
In section \ref{sec:method} we give a brief overview of approximate 
centroiding methods used in our investigation.
In section \ref{sec:data} we discuss the Cram\'{e}r-Rao lower bound satuaration
tests and their corresponding simulated data.
In section \ref{sec:result}, we compare the performances of the methods
discussed in \ref{sec:method}, with the CRLB derived in \ref{sec:data}. Finally,
 we discuss and conclude in \ref{sec:discussion}.               
%%%%%%%%%%%%%%%%%%%%Cram\'{e}r-Rao lower Bound %%%%%%%%%%%%%%

\section{Cram\'{e}r-Rao lower bound on centroiding error}\label{sec:CRLB}

In order to test the accuracy of various centroiding methods, we need to to compare their 
performances against each other at saturating the Cram\'{e}r-Rao lower bound on centroiding error. 
The closer an estimator is to saturating the CRLB, the more information about the quantity that we 
need to estimate is preserved. A measure for closeness of an estimator to 
saturating the CRLB is comparison of the root-mean-squared error of that estimator 
with the CRLB.

The Cram\'{e}r-Rao inequality \citep{cramer} sets a lower bound on the 
root-mean-squared error of unbiased estimators. The CRLB is given by square-root of the inverse of 
the Fisher information matrix $\mathcal{F}$. Thus, in order to find the CRLB,
 it is sufficient to compute the Fisher matrix. Let us assume that there are $M$ observables $\{f_{1}, ... , f_{M}\}$, each
related to $B$ model parameters $\{\theta_{1} , ... , \theta_{B}\}$
\beq
f_{m} = f_{m}(\theta_{1} , ... , \theta_{B}).
\eeq

@@

We examine fitting an exact
 PSF profile to the stars. That is, in our Cram\'{e}r-Rao bound saturation tests, we find the best
 estimates of flux and centroid by optimizing the $\chi^{2}$.
 We expect this method to perform better in determing the centroids
 of stars, and deliver RMSE very close to Cram\'{e}r-Rao bound. In the next section,
 we briefly discuss the centroiding methods used in our Cram\'{e}r-Rao bound saturating tests.

@@

In terms of saturating the Cram\'{e}r-Rao lower bound, we expect this simple 
center of light centroiding to perform worse than all other methods mentioned in 
this section. Hereafter, we call this method $7\times7$ moment centroiding.

@@

An optimal stellar centroiding algorithm must saturate the fundamental Cram\'{e}r-Rao lower 
bound. That is, in all ranges of background noise level, size, radial light profile,
and shape, it must preserve information about the centroids of stars. In practice however,
this is only achievable when we have a reasonably good estimate of the PSF. Since, we do not always 
know the exact PSF profile, we make use of various approximate centroiding algorithms. In this work, we
studied how different methods come close to saturating the CRLB for relatively low signal-to-noise ratio
unsaturated stars.

We focused on examples from two classes of centroiding algorithms. The first class contains fast and approximate
methods that do not require any knowledge of the PSF at the positions of stars. Of methods that belong to this class,
we consider centroiding stars based on fitting a second-order
polynomial to a 3$\times$3 patch of star images smoothed by a Gaussian kernel of fixed width, and finding the center of light 
of a 7$\times$7 patch around the brightest pixel of the star.
The second class of centroiding algorithms contains methods that require knowledge of the PSF (or having a good estimate
of the PSF) at the positions of stars. We considered two examples from this class. The first example
is the matched-filter polynomial centroiding, and the
second example is the full PSF profile fitting. In terms of saturating the Cram\'{e}r-Rao bound, we compared 
the performances of these methods against each other.

@@

On the other hand, the RMSE of centroid estimates of 
default polynomial centroiding are much closer to saturating the CRLB in all ranges of signal-to-noise ratio.
The performance of this method however, is limited by
two important factors: \emph{(i)} signal-to-noise ratio, and \emph{(ii)} PSF size.
The default polynomial centroiding technique only takes advantage of the information
 contained in a 3$\times$3 patch centered on the brighest pixel of the smoothed image which is only
 well-sampled when the FWHM of the simulated image of star
 matches that of the smoothing kernel. Thus, when we apply 
this method to find the centroids of stars with larger FWHM, certain amount of
information (encoded in the Cram\'{e}r-Rao lower bound) is lost, and therefore, the RMSE
of these methods deviates from the CRLB. This deviation becomes even larger at lower
signal-to-noise ratios. Besides, the performance of this method slightly degrades
in the case of undersampled stars (with FWHM close two 2 pixels). 
Presence of noise is another limitting factor.
Although this method is able to get very close to saturating the CRLB in a wide range of
signal-to-noise ratios, it is not reliable in the case of centroiding extremely faint stars.
 This is partly due to the fact that in the presence of noise, 
the brightest pixel of image does not necessarily contain the centroid of stars even after smoothing 
the image.

@@

\bibitem[Bertin \& Arnouts (1996)]{sex} Bertin, E., Arnouts, S. \ 1996  Astronomy and Astrophysics Supplement Series, 117, 393-40
\bibitem[Cram\'{e}r (1946)]{cramer} Cram\'{e}r, H. \ 1946 Mathematical methods of statistics, Princeton university press
\bibitem[Lupton etal. (2001)]{sdss} Lupton, R. etal. \ 2001  arXiv: astro-ph/0101420
\bibitem[Schechter etal. (1993)]{dophot} Schechter, P. L., Mateo, M., Saha, A. \ 1993 \pasp, 1342-1353
\bibitem[Stetson (1987)]{daophot} Stetson, P. B. \ 1987, \pasp, 191-222
\bibitem[Trujillo etal. (2001)]{moffat} Trujillo, I., et al. \ 2001 \mnras, 977-985
